---
title: "GEO 880 - week 4 v2 - Exercise B: Similarity"
author: "Lars Weidinger"
date: "today"
format: 
  html:
    fig_caption: true
    fig-align: center
    highlight: tango
    number-sections: false
    code-fold: true
    theme: lumen
    embed-resources: true
    toc: true
    toc_depth: 2
    toc_float: true
execute:                      # Set global options for execution of code blocks
  echo: true                  # False = no code printed
  warning: false
  message: false
editor: visual
---

<!-- The following html code fragment produces fully justified text alignment -->

```{=html}
<style>
body {text-align: justify}
</style>
```

# Preliminaries

## Set Options

```{r}
options(warning = FALSE)    # Don't show warnings
```

## Install or Load packages

```{r}
#| code-fold: show
library(pacman)

# Load and if not found in the library will attempt to install the package
pacman::p_load(here,
               sf,
               readr,
               tmap,
               dplyr,
               ggplot2,
               tidyr,
               SimilarityMeasures,
               install = TRUE)
```

## Store relevant CRS

```{r}
crs_lv03  <- 21781
crs_lv95  <- 2056
crs_wgs84 <- 4326
```

# Task 1: Similarity measures

## Load & Prepare Data
```{r}
pedestrian <- read_delim("data/pedestrian.csv", ",")

pedestrian<- pedestrian |>
  mutate(Datetime_int = as.integer(DatetimeUTC)) # turn Datetime column from POSIXct to integer
```

# Plot Trajectories
```{r}
ggplot(pedestrian) +
  geom_path(aes(x = E, y = N, colour = TrajID), linewidth = 0.5) +
  geom_point(aes(x = E, y = N, colour = TrajID), size = 0.8) +
  coord_equal() + # preserves the spatial aspect ratio in every subplot
  facet_wrap(~TrajID, ncol = 3) + # creates separate subplots for each trajectory with 3 columns
  labs(
    title = "Six pedestrian trajectories",
    x = "E",
    y = "N",
    colour = "Traj ID"
  ) +
  theme_minimal() # since I prefer this
```

# Task 2: Calculate similarity

## Create indiivudal objects
```{r}
traj1 <- pedestrian |>
  filter(TrajID == 1) |>
  dplyr::select(E, N, Datetime_int) |>
  as.matrix()

traj2 <- pedestrian |>
  filter(TrajID == 2) |>
  dplyr::select(E, N, Datetime_int) |>
  as.matrix()

traj3 <- pedestrian |>
  filter(TrajID == 3) |>
  dplyr::select(E, N, Datetime_int) |>
  as.matrix()

traj4 <- pedestrian |>
  filter(TrajID == 4) |>
  dplyr::select(E, N, Datetime_int) |>
  as.matrix()

traj5 <- pedestrian |>
  filter(TrajID == 5) |>
  dplyr::select(E, N, Datetime_int) |>
  as.matrix()

traj6 <- pedestrian |>
  filter(TrajID == 6) |>
  dplyr::select(E, N, Datetime_int) |>
  as.matrix()
```

I think 2 and 3 are really similar.  
Maybe 2 and 4 are the least similar.  

## Dynamic Time Warp
```{r}
dtw_1_2 <- SimilarityMeasures::DTW(traj1, traj2)

dtw_1_3 <- SimilarityMeasures::DTW(traj1, traj3)

dtw_1_4 <- SimilarityMeasures::DTW(traj1, traj4)

dtw_1_5 <- SimilarityMeasures::DTW(traj1, traj5)

dtw_1_6 <- SimilarityMeasures::DTW(traj1, traj6)
```

## Edit Distance
```{r}
editDist_1_2 <- SimilarityMeasures::EditDist(traj1, traj2, pointDistance = 20)

editDist_1_3 <- SimilarityMeasures::EditDist(traj1, traj3, pointDistance = 20)

editDist_1_4 <- SimilarityMeasures::EditDist(traj1, traj4, pointDistance = 20)

editDist_1_5 <- SimilarityMeasures::EditDist(traj1, traj5, pointDistance = 20)

editDist_1_6 <- SimilarityMeasures::EditDist(traj1, traj6, pointDistance = 20)
```

## Frechet
```{r}
frechet_1_2 <- SimilarityMeasures::Frechet(traj1, traj2)

frechet_1_3 <- SimilarityMeasures::Frechet(traj1, traj3)

frechet_1_4 <- SimilarityMeasures::Frechet(traj1, traj4)

frechet_1_5 <- SimilarityMeasures::Frechet(traj1, traj5)

frechet_1_6 <- SimilarityMeasures::Frechet(traj1, traj6)
```

## Longest Common Subsequence (LCSS)
```{r}
# to make it more efficient we can use a small number in pointSpacing instead of -1 (=infinite)

lcss_1_2 <- SimilarityMeasures::LCSS(traj1, traj2, pointSpacing = 3, pointDistance = 20)

lcss_1_3 <- SimilarityMeasures::LCSS(traj1, traj3, pointSpacing = 3, pointDistance = 20)

lcss_1_4 <- SimilarityMeasures::LCSS(traj1, traj4, pointSpacing = 3, pointDistance = 20)

lcss_1_5 <- SimilarityMeasures::LCSS(traj1, traj5, pointSpacing = 3, pointDistance = 20)

lcss_1_6 <- SimilarityMeasures::LCSS(traj1, traj6, pointSpacing = 3, pointDistance = 20)
```

## Visualisation

### DTW
```{r}
ggplot(
  data.frame(
    comparison = c("2", "3", "4", "5", "6"),
    dtw = c(dtw_1_2, dtw_1_3, dtw_1_4, dtw_1_5, dtw_1_6)
  ),
  aes(x = comparison, y = dtw, fill = comparison)
) +
  geom_col() +
  theme_minimal()
```

### editDist
```{r}
ggplot(
  data.frame(
    comparison = c("2", "3", "4", "5", "6"),
    dtw = c(editDist_1_2, editDist_1_3, editDist_1_4, editDist_1_5, editDist_1_6)
  ),
  aes(x = comparison, y = dtw, fill = comparison)
) +
  geom_col() +
  theme_minimal()
```

### Frechet
```{r}
ggplot(
  data.frame(
    comparison = c("2", "3", "4", "5", "6"),
    dtw = c(frechet_1_2, frechet_1_3, frechet_1_4, frechet_1_5, frechet_1_6)
  ),
  aes(x = comparison, y = dtw, fill = comparison)
) +
  geom_col() +
  theme_minimal()
```

### LCSS
```{r}
ggplot(
  data.frame(
    comparison = c("2", "3", "4", "5", "6"),
    dtw = c(lcss_1_2, lcss_1_3, lcss_1_4, lcss_1_5, lcss_1_6)
  ),
  aes(x = comparison, y = dtw, fill = comparison)
) +
  geom_col() +
  theme_minimal()
```