---
title: "GEO 880 - week 4 v2 - Exercise A: Segmentation"
author: "Lars Weidinger"
date: "today"
format: 
  html:
    fig_caption: true
    fig-align: center
    highlight: tango
    number-sections: false
    code-fold: true
    theme: lumen
    embed-resources: true
    toc: true
    toc_depth: 2
    toc_float: true
execute:                      # Set global options for execution of code blocks
  echo: true                  # False = no code printed
  warning: false
  message: false
editor: visual
---

<!-- The following html code fragment produces fully justified text alignment -->

```{=html}
<style>
body {text-align: justify}
</style>
```

# Preliminaries

## Set Options

```{r}
options(warning = FALSE)    # Don't show warnings
```

## Install or Load packages

```{r}
#| code-fold: show
library(pacman)

# Load and if not found in the library will attempt to install the package
pacman::p_load(here,
               sf,
               readr,
               tmap,
               dplyr,
               ggplot2,
               tidyr,
               install = TRUE)
```

## Store relevant CRS

```{r}
crs_lv03  <- 21781
crs_lv95  <- 2056
crs_wgs84 <- 4326
```

# Load & Prepare Data
```{r}
wildschwein <- read_delim("data/wildschwein_BE_2056.csv", ",")

# Careful! What Timezone is assumed? --> UTC
ruth <- wildschwein |>
    st_as_sf(coords = c("E", "N"), crs = crs_lv95, remove = FALSE) |>
    filter(
      TierName == "Ruth", 
      DatetimeUTC >= "2015-07-01", 
      DatetimeUTC < "2015-07-05"
      )
```

# Task 1: Calculate distances
## a): Specify a temporal window 
In the above dataset, the sampling interval is 15 minutes. If we take a temporal window of 90 minutes, that would mean including 6 fixes. 


## b): Measure the distance to every point within v
```{r}
distance_by_element <- function(later, now) {
  as.numeric(
    st_distance(later, now, by_element = TRUE)
  )
}

ruth <- ruth |>
    mutate(
      nMinus3 = distance_by_element(lag(geometry, 3), geometry),  # distance to pos -45 min
        nMinus2 = distance_by_element(lag(geometry, 2), geometry),  # distance to pos -30 min
        nMinus1 = distance_by_element(lag(geometry, 1), geometry),  # distance to pos -15 min
        nPlus1  = distance_by_element(geometry, lead(geometry, 1)), # distance to pos +15 min
        nPlus2  = distance_by_element(geometry, lead(geometry, 2)),  # distance to pos +30 min
      nPlus3  = distance_by_element(geometry, lead(geometry, 3))  # distance to pos +45 min
    )
```

```{r}
# calculate mean distance
ruth <- ruth |>
    rowwise() |>
    mutate(
        stepMean = mean(c(nMinus2, nMinus1, nPlus1, nPlus2))
    ) |>
    ungroup()

ruth
```

# Task 2: Specify and apply threshold d
```{r}
summary <- summary(ruth$stepMean)

ruth_threshold <- mean(ruth$stepMean, na.rm = TRUE) # do not use the NA values
```

```{r}
# use the mean step length as the mean
ruth <- ruth |>
    mutate(static = stepMean < mean(stepMean, na.rm = TRUE))
```

# Task 3: Visualize segmented trajectories
```{r}
ggplot(ruth) +
  geom_path(aes(x = E, y = N, colour = static), linewidth = 0.5) +
  geom_point(aes(x = E, y = N, colour = static), size = 0.8) +
  coord_equal() +
  scale_colour_manual(
    values = c("TRUE" = "purple", "FALSE" = "orange"),
    labels = c("TRUE" = "Static", "FALSE" = "Movement")
  ) +
  labs(
    title = "Segmented trajectory of Ruth",
    x = "E",
    y = "N",
    colour = "Behaviour"
  ) +
  theme_minimal() # since I prefer this
```

# Task 4: Segment-based analysis
```{r}
# function to create unique IDs
rle_id <- function(vec) {
    x <- rle(vec)$lengths
    as.factor(rep(seq_along(x), times = x))
}
```

```{r}
# create unique IDs
ruth <- ruth |>
    mutate(segment_id = rle_id(static))
```

```{r}
ggplot(ruth) +
  geom_path(aes(x = E, y = N, colour = segment_id, group = segment_id), linewidth = 0.5) +
  geom_point(aes(x = E, y = N, colour = segment_id), size = 0.8) +
  coord_equal() +
  labs(
    title = "Trajectory segments of Ruth",
    x = "E",
    y = "N",
    colour = "Segment ID"
  ) +
  theme_minimal() # since I prefer this
```

```{r}
# calculate duration for each segment
segment_duration <- ruth |>
  group_by(segment_id, static) |>
  summarise(
    start_time = min(DatetimeUTC),
    end_time   = max(DatetimeUTC),
    duration_mins = as.numeric(
      difftime(end_time, start_time, units = "mins")
    ),
    n = n(),
    .groups = "drop"
  )
```

```{r}
# remove short segments, keep only segments 5+ minutes
long_segments <- segment_duration |>
  filter(duration_mins >= 5)

ruth_filtered <- ruth |>
  filter(segment_id %in% long_segments$segment_id)
```


